On the minimum leaf conductance: its role in models of plant water use, and ecological and environmental controls

### 2018-07-10 manuscript.Rmd
Final revisions

During conditions where photosynthesis does not take place, such as severe drought, extreme temperature or low light, it seems advantageous for plants to close stomata and completely halt water loss. However, water loss continues through the cuticle and incompletely closed stomata, together constituting the leaf minimum conductance (g~min~). In this review, we critically evaluate sources of variation in g~min~, quantitatively compare various methods for its estimation, and illustrate the role of g~min~ in models of leaf gas exchange. A literature compilation of g~min~ as measured by weight loss of detached leaves is presented, which shows much variation in this trait that is not clearly related to species groups, climate of origin, or leaf type. Much evidence points to the idea that g~min~ is highly responsive to the growing conditions of the plant, including soil water availability, temperature, and air humidity - as we further demonstrate with two case studies. We pay special attention to the role of the minimum conductance in the Ball-Berry model of stomatal conductance, and caution against the usual regression-based method for its estimation. The synthesis presented here provides guidelines for use of g~min~ in ecosystem models, and points to clear research gaps for this drought tolerance trait.

@@

\
Plants face a dilemma in constructing leaves that minimize water loss while allowing the uptake of CO~2~. Since a membrane that is permeable to CO~2~ but not H~2~O has never evolved, all land plants have stomata in their leaves, which disrupt the cuticle and allow CO~2~ uptake. It is well known that stomata open and close in response to changes in light intensity, humidity and CO~2~ concentration at the leaf surface. Stomata can be thought to balance the uptake of CO~2~ with the loss of water, and thus to close stomata when photosynthetic rates are low. To avoid desiccation and ultimate death, stomata typically close during periods of severe water stress. When stomata are closed, water loss continues at a greatly diminished rate through the cuticle. After accounting for evaporative demand, this rate of water loss is expressed as the minimum conductance of a leaf. There is increasing recognition that the minimum conductance plays an important role in estimating water fluxes in plant canopies [@barnard2013], during heat waves [@kala2016], and in models of plant drought response [@blackman2016; @martinstpaul2017].

@@

# II. Comparison of various definitions of minimum conductance

\
Measurements of minimum conductance, after stomatal closure is either induced or assumed, can be broadly divided into the following categories. The absolute minimum conductance attained by leaves is that through the cuticle only. Measurements of cuticular conductance (g~cuti~) are typically made on isolated non-stomatal cuticles [@riederer2001], via special gas exchange techniques [@boyer1997], or mass loss of detached leaves by sealing the side with stomata [@kerstiens1996]. We further discuss some important aspects of cuticular conductance in Section III.

@@

\
We contrasted g~cuti~ and g~min~ with three estimates of stomatal conductance when photosynthesis rates are low or zero during non-drought conditions: night-time conductance (g~dark~) (further discussed in Section VI.4), conductance at low photosynthetically active radiation (PAR) (0 - 40 $\mu$ mol m^-2^ s^-1^) during the day, and conductance during conditions when photosynthesis rates are very low (< 1 $\mu$ mol m^-2^ s^-1^) (but excluding low PAR and drought).

\
We quantitatively compared each of the different conductances by synthesizing existing data. Measurements of conductance of isolated, non-stomatal cuticles (g~cuti~) were taken from the compilations by @kerstiens1996 and @schuster2017 (both also include g~min~ data but these were not used since they included no metadata). Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see Methods S1 for description and references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Methods S1 for full details of the compilation). The database includes a total of 221 species (136 Angiosperm, 49 Gymnosperm, 1 Pteridophyte), from 57 taxonomic families, with woody species making up roughly 2/3 of the data.

@@

\
Although we use the term cuticular conductance freely, this transport pathway does not represent a true conductance since water does not diffuse as a gas through the cuticle. Instead, it dissolves into the medium of the cuticle, diffuses through the solid matrix, and is desorped at the outer edge of the cuticle [@kerstiens1996; @riederer2001; @schreiber1996]. The main barrier to diffusion is actually a very thin layer of wax at the leaf surface. Because most of the resistance is located in such a thin layer, g~cuti~ does not correlate with the thickness of the cuticle [@priestley1943; @riederer2001; @anfodillo2002; @schuster2016b]. Nonetheless there is considerable variation in cuticle thickness among plant species [@schuster2016a], along altitudinal transects [@delucia1984], and even with increasing height in the canopy of very tall trees [@woodruff2010]. If the cuticle thickness does not directly affect its conductance, what is the value of increasing the cuticle?

@@

The minimum leaf conductance includes two pathways: across the cuticle, and through the (potentially incompletely closed) stomata. Only a few studies have directly quantified the stomatal component of minimum conductance, in contrast to the wealth of information on the cuticular component. A detailed study of *Hedera helix*  concluded that 35% of water loss occurred across the stomatal pores (though closed), and 65% across the non-stomatal part of the cuticle, despite the fact that the stomata presumably covered only a small fraction of the leaf [@santrucek2004]. For ten *Sorghum* genotypes, @muchow1989 reported a strong positive correlation between stomatal density and g~min~, suggesting leaky stomata contribute substantially to g~min~. For seven conifer species, @brodribb2014 reported that 50 - 94% of water loss of detached leaves originated from the stomatal side, concluding that stomata must be very leaky, and likely incompletely closed. However, @santrucek2004 found that the cuticle of the stomatal side (but excluding stomata itself) in *Hedera helix* was many times more permeable than the non-stomatal side, thus providing another potential explanation for this difference. Either way, water loss rates of intact leaves cannot be simply explained from the permeability of isolated non-stomatal cuticles, and the contribution of leaky stomata, and potentially more permeable cuticle on the abaxial side, likely varies among species.

@@

## 1. Minimum conductance shows little meaningful interspecific variation

@@

\
A number of studies have attempted to attribute the drought-induced acclimation in g~min~ to a change in the chemical composition of the cuticle. @bengtson1978 reported lower g~min~ in six oat varieties in response to drought, and an increase in the amount of cuticular waxes, but could not find a relationship between the two and the response was highly genotype-specific. However, @premachandra1992 found that epicuticular wax load increased on leaves of non-irrigated Sorghum cultivars and was positively correlated with cuticular conductance and cell membrane stability. @mackova2013 found that adding ABA (simulating drought stress) to *Lepidium sativum* increased the chain length of cuticular waxes (but not total amounts). @bi2017 also reported that drought caused a change in cuticular wax production and chemical composition, but again in a highly genotype-specific way. Thus it can be concluded that the cuticle indeed changes substantially after a change in plant available water, but in a complex, species-specific manner that is yet to be connected directly to changes in g~min~.

@@

\
A number of studies have assumed that the g~0~ in the Ball-Berry model equals the night-time conductance (g~dark~) [@barnard2013; @lombardozzi2017], simply because it is a condition when net photosynthesis is zero. We provide some caution to the assumption that g~dark~ can be used in models of daytime leaf conductance. First, ample observations suggest that g~dark~ is not a fixed rate but varies tremendously during the night. @caird2007 (and references therein) described how for many species, g~dark~ is not stable throughout the night period. Instead, endogenous, gradual increases in stomatal opening during predawn hours have been reported in many species under natural field conditions as well as in controlled environments [@rawson1988]. @rescodedios2016 showed that, in *Eucalyptus camaldulensis*, g~dark~ in the period just after sunset was much lower than pre-dawn g~dark~. Strong evidence for endogenous regulation of g~dark~ was reported by @rescodedios2013, who showed that g~dark~ fluctuated throughout the night despite environmental conditions being held constant in whole-tree chambers. In addition, g~dark~ can show a clear response to vapour pressure deficit [@barbour2007] during night-time conditions. Clearly g~dark~ is an actively controlled process that cannot be adequately summarized by a single constant g~0~, and should be modelled in a separate framework that is yet to be identified.

>>>>

When the rate of photosynthesis is greatly diminished, such as during severe drought, extreme temperature or low light, it seems advantageous for plants to close stomata and completely halt water loss. However, water loss continues through the cuticle and incompletely closed stomata, together constituting the leaf minimum conductance (g~min~). In this review, we critically evaluate sources of variation in g~min~, quantitatively compare various methods for its estimation, and illustrate the role of g~min~ in models of leaf gas exchange. A literature compilation of g~min~ as measured by weight loss of detached leaves is presented, which shows much variation in this trait that is not clearly related to species groups, climate of origin, or leaf type. Much evidence points to the idea that g~min~ is highly responsive to the growing conditions of the plant, including soil water availability, temperature, and air humidity - as we further demonstrate with two case studies. We pay special attention to the role of the minimum conductance in the Ball-Berry model of stomatal conductance, and caution against the usual regression-based method for its estimation. The synthesis presented here provides guidelines for use of g~min~ in ecosystem models, and points to clear research gaps for this drought tolerance trait.

@@

\
Plants face a dilemma in constructing leaves that minimize water loss while allowing the uptake of CO~2~. Since a membrane that is permeable to CO~2~ but not H~2~O has never evolved, all land plants have stomata in their leaves, which disrupt the cuticle and allow CO~2~ uptake. It is well known that stomata open and close in response to changes in light intensity, humidity and CO~2~ concentration at the leaf surface. To avoid desiccation and ultimate death, stomata typically close during periods of water stress. When stomata are closed, water loss continues at a greatly diminished rate through the cuticle. After accounting for evaporative demand, this rate of water loss is expressed as the minimum conductance of a leaf. There is increasing recognition that the minimum conductance plays an important role in estimating water fluxes in plant canopies [@barnard2013], during heat waves [@kala2016], and in models of plant drought response [@blackman2016; @martinstpaul2017].

@@

# II. Comparison of various definitions and measurement techniques of minimum conductance

\
Measurements of minimum conductance, after stomatal closure is either induced or assumed, can be broadly divided into the following categories: conductance of the cuticle only, conductance of detached leaves, and gas-exchange measurements during conditions leading to presumed stomatal closure. The absolute minimum conductance attained by leaves is that through the cuticle only. Measurements of cuticular conductance (g~cuti~) are typically made on isolated non-stomatal cuticles [@riederer2001], via special gas exchange techniques [@boyer1997], or mass loss of detached leaves by sealing the side with stomata [@kerstiens1996]. We further discuss some important aspects of cuticular conductance in Section III.

@@

\
We quantitatively compared g~cuti~ and g~min~ by synthesizing existing data. In addition, we compiled data on leaf conductance when photosynthesis rates are low or zero during non-drought conditions: night-time conductance (g~dark~) (further discussed in Section VI.4), conductance at low photosynthetically active radiation (PAR) (0 - 40 $\mu$ mol m^-2^ s^-1^) during the day, and conductance during conditions when photosynthesis rates are very low (< 1 $\mu$ mol m^-2^ s^-1^) (but excluding low PAR and drought).

\
Measurements of conductance of isolated, non-stomatal cuticles (g~cuti~) were taken from the compilations by @kerstiens1996 and @schuster2017 (both also include g~min~ data but these were not used since they included no metadata). Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see Methods S1 for description and references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Methods S1 for full details of the compilation). The database includes a total of 221 species (136 Angiosperm, 49 Gymnosperm, 1 Pteridophyte), from 57 taxonomic families, with woody species making up roughly 2/3 of the data.

@@

\
Although we use the term cuticular conductance freely, this transport pathway does not represent a true conductance since water does not diffuse as a gas through the cuticle. Instead, it dissolves into the medium of the cuticle, diffuses through the solid matrix, and is desorped at the outer edge of the cuticle [@kerstiens1996; @riederer2001; @schreiber1996]. The main barrier to diffusion is actually a very thin layer of wax at the leaf surface. Because most of the resistance is located in such a thin layer, g~cuti~ does not correlate with the thickness of the cuticle [@priestley1943; @riederer2001; @anfodillo2002; @schuster2016b]. Nonetheless there is considerable variation in cuticle thickness among plant species [@schuster2016a], along altitudinal transects [@delucia1984], and even with increasing height in the canopy of very tall trees [@woodruff2010]. If the cuticle thickness does not directly affect its conductance, what is the value of a thicker cuticle?

@@

\
The minimum leaf conductance includes two pathways: across the cuticle, and through the (potentially incompletely closed) stomata. Only a few studies have directly quantified the stomatal component of minimum conductance, in contrast to the wealth of information on the cuticular component. A detailed study of *Hedera helix*  concluded that 35% of water loss occurred across the stomatal pores (though closed), and 65% across the non-stomatal part of the cuticle, despite the fact that the stomata presumably covered only a small fraction of the leaf [@santrucek2004]. For ten *Sorghum* genotypes, @muchow1989 reported a strong positive correlation between stomatal density and g~min~, suggesting leaky stomata contribute substantially to g~min~. For seven conifer species, @brodribb2014 reported that 50 - 94% of water loss of detached leaves originated from the stomatal side, concluding that stomata must be very leaky, and likely incompletely closed. However, @santrucek2004 found that the cuticle of the stomatal side (but excluding stomata itself) in *Hedera helix* was many times more permeable than the non-stomatal side, thus providing another potential explanation for this difference. Either way, water loss rates of intact leaves cannot be simply explained from the permeability of isolated non-stomatal cuticles, and the contribution of leaky stomata, and potentially more permeable cuticle on the abaxial side, likely varies among species.

@@

## 1. Minimum conductance is highly variable among species

@@

\
A number of studies have attempted to attribute the drought-induced acclimation in g~min~ to a change in the chemical composition of the cuticle. @bengtson1978 reported lower g~min~ in six oat varieties in response to drought, and an increase in the amount of cuticular waxes, but could not find a relationship between the two and the response was highly genotype-specific. However, @premachandra1992 found that epicuticular wax load increased on leaves of non-irrigated *Sorghum* cultivars and was positively correlated with cuticular conductance and cell membrane stability. @mackova2013 found that adding ABA (simulating drought stress) to *Lepidium sativum* increased the chain length of cuticular waxes (but not total amounts). @bi2017 also reported that drought caused a change in cuticular wax production and chemical composition, but again in a highly genotype-specific way. Thus it can be concluded that the cuticle indeed changes substantially after a change in plant available water, but in a complex, species-specific manner that is yet to be connected directly to changes in g~min~.

@@

\
A number of studies have assumed that the g~0~ in the Ball-Berry model equals the night-time conductance (g~dark~) [@barnard2013; @lombardozzi2017], simply because it is a condition when net photosynthesis is zero. We provide some caution to the assumption that g~dark~ can be used in models of daytime leaf conductance. First, ample observations suggest that g~dark~ is not a fixed rate but varies tremendously during the night. @caird2007 (and references therein) described how for many species, g~dark~ is not stable throughout the night period. Instead, endogenous, gradual increases in stomatal opening during predawn hours have been reported in many species under natural field conditions as well as in controlled environments [@rawson1988]. @rescodedios2016 showed that, in *Eucalyptus camaldulensis*, g~dark~ in the period just after sunset was much lower than pre-dawn g~dark~. Strong evidence for endogenous regulation of g~dark~ was reported by @rescodedios2013, who showed that g~dark~ fluctuated throughout the night despite environmental conditions being held constant in whole-tree chambers. In addition, g~dark~ can show a clear response to vapour pressure deficit [@barbour2007] during night-time conditions. Finally, g~dark~ is under strong genetic control independent of day-time water use - at least in grapevine [@coupelledru2016]. Clearly g~dark~ is an actively controlled process that cannot be adequately summarized by a single constant g~0~, and should be modelled in a separate framework that is yet to be identified.
